\documentclass{article}
\usepackage{graphicx, nips} % Required for inserting images
\usepackage{listings}
\usepackage{xcolor}
\usepackage{url}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage{multicol}
\usepackage[compact]{titlesec}

% Reduce spacing
\setlength{\parskip}{2pt}
\setlength{\parsep}{0pt}
\setlength{\headsep}{10pt}
\setlength{\topskip}{0pt}
\setlength{\topmargin}{-10pt}
\setlength{\topsep}{0pt}
\setlength{\partopsep}{0pt}
\setlength{\itemsep}{1pt}

% Compact sections
\titlespacing{\section}{0pt}{6pt plus 2pt minus 2pt}{3pt plus 2pt minus 2pt}
\titlespacing{\subsection}{0pt}{4pt plus 2pt minus 2pt}{2pt plus 2pt minus 2pt}
\titlespacing{\subsubsection}{0pt}{3pt plus 2pt minus 2pt}{1pt plus 2pt minus 2pt}

% SQL code style - compact
\lstdefinestyle{sqlstyle}{
    language=SQL,
    basicstyle=\ttfamily\scriptsize,
    keywordstyle=\color{blue}\bfseries,
    commentstyle=\color{gray},
    stringstyle=\color{red},
    showstringspaces=false,
    breaklines=true,
    frame=none,
    numbers=none,
    aboveskip=2pt,
    belowskip=2pt
}

\title{CSC3170 Project Report: Dormitory Management System}
\author{Xing Qiliang - 123090669}

\begin{document}
\maketitle

\section{Project Description}


\subsection{System Overview}
This project implements a Dormitory Management System using modern web architecture with strict frontend-backend separation. The backend uses FastAPI (Python) with MySQL database providing RESTful APIs, while the frontend uses Vue.js 3 with Element Plus UI framework, communicating via HTTP/HTTPS with JWT authentication. The system supports two roles (Student and Administrator) with 21+ functional modules covering dormitory management, student registration, room assignment, maintenance requests, bill management, and approval workflows.


\section{Requirements Analysis}
\subsection{Student Role (8 Functions)}
(1) Authentication with Argon2 password hashing; (2) Profile management and password change; (3) View dormitory details and roommates; (4) View bills with payment status; (5) Submit or modify room change requests; (6) Submit or modify maintenance requests with priority; (7) Track request status; (8) Dashboard overview.

\subsection{Administrator Role (12+ Functions)}
(1) CRUD operations on students; (2) Dormitory management with occupancy stats; (3) Room assignment/reassignment; (4) View resident information; (5) Approve/reject room change requests; (6) Process maintenance requests; (7) Bill management; (8) Statistical dashboard with system-wide metrics; (9) Profile management; (10) Batch operations with pagination; (11) Monitor student distribution.

The system implements 21+ distinct operations exceeding the 9-function requirement, with all database operations handled through backend APIs ensuring security and integrity.

\section{Database and SQL Design}
\subsection{Database Schema and ER Model}
The system uses 6 normalized tables following Third Normal Form (3NF) principles:

\begin{table}[htbp]
\centering
\scriptsize
\begin{minipage}{0.43\textwidth}
\centering
\begin{tabular}{@{}p{1.8cm}p{1.4cm}p{1.2cm}p{1.6cm}@{}}
\toprule
\textbf{Field} & \textbf{Type} & \textbf{Constraint} & \textbf{Description} \\ \midrule
\multicolumn{4}{c}{\textbf{Students Table}} \\ \midrule
student\_id & VARCHAR(20) & PK & Student ID \\
password & VARCHAR(255) & NOT NULL & Hashed pwd \\
name & VARCHAR(100) & NOT NULL & Full name \\
gender & ENUM & NOT NULL & Male/Female \\
nationality & VARCHAR(50) & NOT NULL & Nationality \\
college & VARCHAR(50) & NOT NULL & College \\
enrollment\_year & INT & NOT NULL & Enroll year \\
email & VARCHAR(100) & UNIQUE & Email \\
dorm\_id & INT & FK (null) & Dorm ID \\
created\_at, updated\_at & DATETIME & AUTO & Timestamps \\ \midrule
\multicolumn{4}{l}{\textit{Rel: Many-to-One with Dormitories}} \\ \bottomrule
\end{tabular}
\end{minipage}%
\hspace{0.14\textwidth}%
\begin{minipage}{0.43\textwidth}
\centering
\begin{tabular}{@{}p{1.8cm}p{1.4cm}p{1.2cm}p{1.6cm}@{}}
\toprule
\textbf{Field} & \textbf{Type} & \textbf{Constraint} & \textbf{Description} \\ \midrule
\multicolumn{4}{c}{\textbf{Dormitories Table}} \\ \midrule
dorm\_id & INT & PK, AUTO & Dorm ID \\
building\_no & VARCHAR(10) & NOT NULL & Building \\
floor\_no & INT & NOT NULL & Floor \\
room\_no & VARCHAR(20) & UNIQUE & Room No. \\
gender\_type & ENUM & NOT NULL & Male/Female \\
total\_beds & INT & DEFAULT 4 & Capacity \\
occupied\_beds & INT & DEFAULT 0 & Occupancy \\
created\_at, updated\_at & DATETIME & AUTO & Timestamps \\ \midrule
\multicolumn{4}{l}{\textit{Rel: One-to-Many with Students, Bills}} \\ \bottomrule
\end{tabular}
\end{minipage}
\end{table}

\begin{table}[htbp]
\centering
\scriptsize
\begin{minipage}{0.43\textwidth}
\centering
\begin{tabular}{@{}p{1.8cm}p{1.4cm}p{1.2cm}p{1.6cm}@{}}
\toprule
\textbf{Field} & \textbf{Type} & \textbf{Constraint} & \textbf{Description} \\ \midrule
\multicolumn{4}{c}{\textbf{Dorm Change Requests}} \\ \midrule
request\_id & INT & PK, AUTO & Request ID \\
student\_id & VARCHAR(20) & FK & Student \\
current\_dorm\_id & INT & FK & Current dorm \\
target\_dorm\_id & INT & FK & Target dorm \\
reason & TEXT & & Reason \\
status & ENUM & pending & Status \\
admin\_id & INT & FK (null) & Admin \\
admin\_comment & TEXT & & Comment \\
created\_at, updated\_at & DATETIME & AUTO & Timestamps \\ \midrule
\multicolumn{4}{l}{\textit{Rel: M-to-1 with Students, Admins, Dorms}} \\ \bottomrule
\end{tabular}
\end{minipage}%
\hspace{0.14\textwidth}%
\begin{minipage}{0.43\textwidth}
\centering
\begin{tabular}{@{}p{1.8cm}p{1.4cm}p{1.2cm}p{1.6cm}@{}}
\toprule
\textbf{Field} & \textbf{Type} & \textbf{Constraint} & \textbf{Description} \\ \midrule
\multicolumn{4}{c}{\textbf{Maintenance Requests}} \\ \midrule
request\_id & INT & PK, AUTO & Request ID \\
student\_id & VARCHAR(20) & FK & Student \\
dorm\_id & INT & FK & Dormitory \\
issue\_type & VARCHAR(50) & NOT NULL & Issue type \\
description & TEXT & NOT NULL & Details \\
status & ENUM & pending & Status \\
priority & ENUM & medium & Priority \\
admin\_id & INT & FK (null) & Admin \\
admin\_comment & TEXT & & Comment \\
completed\_at & DATETIME & & Completed \\
created\_at, updated\_at & DATETIME & AUTO & Timestamps \\ \midrule
\multicolumn{4}{l}{\textit{Rel: M-to-1 with Students, Dorms, Admins}} \\ \bottomrule
\end{tabular}
\end{minipage}
\end{table}

\begin{table}[htbp]
\centering
\scriptsize
\begin{minipage}{0.43\textwidth}
\centering
\begin{tabular}{@{}p{1.8cm}p{1.4cm}p{1.2cm}p{1.6cm}@{}}
\toprule
\textbf{Field} & \textbf{Type} & \textbf{Constraint} & \textbf{Description} \\ \midrule
\multicolumn{4}{c}{\textbf{Bills Table}} \\ \midrule
bill\_id & INT & PK, AUTO & Bill ID \\
dorm\_id & INT & FK & Dormitory \\
bill\_type & VARCHAR(50) & NOT NULL & Type \\
amount & DECIMAL(10,2) & NOT NULL & Amount \\
billing\_month & VARCHAR(7) & NOT NULL & YYYY-MM \\
due\_date & DATE & NOT NULL & Due date \\
status & ENUM & unpaid & Status \\
paid\_at & DATETIME & & Paid time \\
created\_at, updated\_at & DATETIME & AUTO & Timestamps \\ \midrule
\multicolumn{4}{l}{\textit{Rel: Many-to-One with Dormitories}} \\ \bottomrule
\end{tabular}
\end{minipage}%
\hspace{0.14\textwidth}%
\begin{minipage}{0.43\textwidth}
\centering
\begin{tabular}{@{}p{1.8cm}p{1.4cm}p{1.2cm}p{1.6cm}@{}}
\toprule
\textbf{Field} & \textbf{Type} & \textbf{Constraint} & \textbf{Description} \\ \midrule
\multicolumn{4}{c}{\textbf{Administrators Table}} \\ \midrule
admin\_id & INT & PK, AUTO & Admin ID \\
username & VARCHAR(50) & UNIQUE & Username \\
password & VARCHAR(255) & NOT NULL & Hashed pwd \\
name & VARCHAR(100) & NOT NULL & Full name \\
email & VARCHAR(100) & UNIQUE & Email \\
role & ENUM & admin & Role \\
phone & VARCHAR(20) & & Phone \\
is\_active & BOOLEAN & TRUE & Active \\
last\_login & DATETIME & & Last login \\
created\_at, updated\_at & DATETIME & AUTO & Timestamps \\ \midrule
\multicolumn{4}{l}{\textit{Rel: 1-to-M with DormChange, Maintenance}} \\ \bottomrule
\end{tabular}
\end{minipage}
\end{table}

\subsection{Normalization and Constraints}
The schema eliminates redundancy through proper normalization: student information is not duplicated in request tables (using foreign keys instead), dormitory details are centralized, and all relationships use referential integrity constraints. Cascade rules ensure data consistency: deleting a student cascades to their requests (ON DELETE CASCADE), while deleting a dormitory sets student dorm\_id to NULL (ON DELETE SET NULL) to preserve historical records.

\subsection{Key SQL Examples}
\subsubsection{Dormitory Search with Filters}
\begin{lstlisting}[style=sqlstyle]
SELECT d.dorm_id, d.building_no, d.floor_no, d.room_no, d.gender_type, 
       d.total_beds, d.occupied_beds, (d.total_beds - d.occupied_beds) AS available_beds
FROM dormitories d
WHERE d.gender_type = ? AND (d.total_beds - d.occupied_beds) > 0
  AND (d.building_no = ? OR ? IS NULL)
ORDER BY d.building_no, d.floor_no, d.room_no LIMIT ? OFFSET ?;
\end{lstlisting}

\subsubsection{Statistical Dashboard Query}
\begin{lstlisting}[style=sqlstyle]
SELECT COUNT(DISTINCT s.student_id) AS total_students,
       COUNT(DISTINCT d.dorm_id) AS total_dorms,
       SUM(d.total_beds - d.occupied_beds) AS available_beds,
       COUNT(CASE WHEN b.status = 'unpaid' THEN 1 END) AS unpaid_bills,
       COUNT(CASE WHEN dr.status = 'pending' THEN 1 END) AS pending_changes,
       COUNT(CASE WHEN s.gender = 'Male' THEN 1 END) AS male_students
FROM students s
LEFT JOIN dormitories d ON s.dorm_id = d.dorm_id
LEFT JOIN bills b ON d.dorm_id = b.dorm_id
LEFT JOIN dorm_change_requests dr ON s.student_id = dr.student_id;
\end{lstlisting}

\subsubsection{Room Change Approval Transaction}
\begin{lstlisting}[style=sqlstyle]
START TRANSACTION;
UPDATE dorm_change_requests SET status = 'approved', admin_id = ?, 
  admin_comment = ? WHERE request_id = ? AND status = 'pending';
UPDATE students SET dorm_id = (SELECT target_dorm_id FROM dorm_change_requests 
  WHERE request_id = ?) WHERE student_id = (SELECT student_id FROM dorm_change_requests WHERE request_id = ?);
UPDATE dormitories SET occupied_beds = occupied_beds - 1 
  WHERE dorm_id = (SELECT current_dorm_id FROM dorm_change_requests WHERE request_id = ?);
UPDATE dormitories SET occupied_beds = occupied_beds + 1 
  WHERE dorm_id = (SELECT target_dorm_id FROM dorm_change_requests WHERE request_id = ?);
COMMIT;
\end{lstlisting}

\section{Frontend and Backend Implementation}
\subsection{Backend Architecture (FastAPI + MySQL)}
\textbf{Technology Stack:} FastAPI 0.104.1 (async web framework), SQLAlchemy 2.0.23 (ORM), PyMySQL 1.1.0 (MySQL driver), MySQL 8.0+ (database), python-jose 3.3.0 (JWT), argon2-cffi 23.1.0 (password hashing), Pydantic 2.5.0 (validation), python-dotenv 1.0.0 (configuration).

\textbf{Three-Layer Architecture:}

(1) \textit{Data Layer (models.py):} SQLAlchemy ORM models define database schema with declarative base. Each model includes relationships using \texttt{relationship()} and \texttt{back\_populates}, cascade rules (CASCADE for dependent data, SET NULL for optional references), and automatic timestamp tracking. Enum types ensure data consistency for status fields.

(2) \textit{Business Logic Layer (routers):} Three router modules implement RESTful endpoints:
\begin{itemize}
\item \texttt{auth.py} (4 endpoints): POST /login (authenticate and issue JWT), POST /register (create new student account), GET /me (retrieve current user info), POST /logout (invalidate token).
\item \texttt{students.py} (12 endpoints): Profile management (GET/PUT /profile, PUT /password), dormitory operations (GET /dormitory, /roommates, /dormitories with filters), financial (GET /bills with status filter), requests (GET/POST /dorm-change, GET/POST /maintenance, PUT /maintenance/\{id\}).
\item \texttt{admin.py} (21 endpoints): Student management (GET/GET/PUT/DELETE /students), dormitory management (GET/GET/PUT /dormitories, GET /dormitories/\{id\}/students), request processing (GET/PUT /dorm-change, POST /dorm-change/\{id\}/approve, POST /dorm-change/\{id\}/reject, GET/PUT /maintenance), billing (GET/POST/PUT/DELETE /bills), statistics (GET /statistics), profile (PUT /profile, PUT /password).
\end{itemize}

(3) \textit{Validation Layer (schemas.py):} Pydantic models enforce type safety and validation rules. Examples: student\_id must be exactly 9 characters, passwords minimum 6 characters with , emails validated with EmailStr type, enrollment\_year constrained to realistic range (2020-2030), amounts positive with 2 decimal precision.

\textbf{Security Implementation:}

\textit{Password Security:} Argon2id algorithm with parameters memory\_cost=65536 KB (64MB), time\_cost=3 iterations, parallelism=4 threads. Salt automatically generated per password.

\textit{Authentication Flow:} (1) User submits credentials; (2) Backend queries database and verifies password hash; (3) On success, generates JWT with payload \{user\_id, user\_type, exp\}; (4) Client stores token in localStorage; (5) Subsequent requests include token in Authorization header (Bearer scheme); (6) Protected endpoints use dependency injection (\texttt{Depends(get\_current\_student)}) to verify token and extract user identity.

\textit{Authorization:} Role-based access control ensures students can only access their own data (checked by comparing token user\_id with resource owner), while administrators have elevated privileges verified through user\_type claim in JWT payload.

\subsection{Frontend Architecture (Vue.js 3)}
\textbf{Technology Stack:} Vue.js 3.4.0 (Composition API with \texttt{<script setup>} syntax), Element Plus 2.5.0 (UI components), Vue Router 4.2.5 (SPA routing), Axios 1.6.2 (HTTP client), Vite 5.0.8 (build tool with HMR), ECharts 5.4.3 (data visualization).

\textbf{Component Architecture:}

\textit{Layout Components:} Two master layouts provide consistent navigation and user experience. \texttt{AdminLayout.vue} includes top navigation bar with system name, user info dropdown (profile/logout), and side menu (Dashboard, Students, Dormitories, Requests, Maintenance, Bills). \texttt{StudentLayout.vue} features simplified navigation focused on student-specific functions.

\textit{View Components (15 pages):} Student interface includes Dashboard , Profile , Dormitory , Bills , DormChange , Maintenance . Admin interface includes enhanced Dashboard , Students management , Dormitories , DormRequests , Maintenance processing , Bills management .

\textit{API Service Layer:} Centralized API functions in \texttt{/api} directory with modules for auth, student, admin operations. Each module exports typed async functions that handle request construction, error catching, and response parsing.

\textbf{State Management and Routing:}

Vue Router implements protected routes with \texttt{beforeEach} navigation guard checking localStorage for valid JWT token and user type. Unauthorized access redirects to login page. Authentication state managed via composable utilities (\texttt{useAuth.js}) providing reactive refs for \texttt{isLoggedIn}, \texttt{userType}, \texttt{currentUser}.

\textbf{User Experience Features:}

Form validation provides instant feedback using Element Plus validation rules (required fields, email format, numeric ranges, custom validators). Loading states display skeleton screens during data fetching. Error handling shows user-friendly toast messages parsed from backend error responses. Confirmation dialogs prevent accidental deletions or rejections. Search and filter functionality includes debouncing (300ms delay) to reduce API calls. Responsive design adapts to mobile/tablet/desktop using CSS media queries and Element Plus responsive grid system. Pagination with customizable page sizes (10/20/50 items) handles large datasets efficiently.

\subsection{Communication Protocol and Integration}
\textbf{RESTful API Design:} All endpoints follow REST principles with semantic HTTP methods (GET for retrieval, POST for creation, PUT for updates, DELETE for removal) and resource-based URLs (e.g., \texttt{/api/students/\{id\}} for individual student). Request/response bodies use JSON format with camelCase naming convention.

\textbf{Request Flow:} (1) User action triggers frontend function; (2) Axios interceptor adds JWT token to Authorization header (\texttt{Bearer <token>}); (3) Request sent to backend with JSON payload; (4) Backend validates token, checks permissions, processes request; (5) Response returned with HTTP status code and JSON body; (6) Axios response interceptor catches errors globally; (7) Frontend updates UI with success message or displays error notification.

\textbf{Error Handling Strategy:} Backend returns consistent error format: \texttt{\{"detail": "Error message"\}} with appropriate HTTP status (400 for validation errors, 401 for authentication failures, 403 for authorization denials, 404 for not found, 500 for server errors). Frontend interceptor maps status codes to user-friendly messages and displays toast notifications. Token expiration (401) triggers automatic redirect to login page.

\textbf{CORS Configuration:} Backend enables Cross-Origin Resource Sharing to allow frontend (running on different port during development) to access API. Configured with \texttt{allow\_origins=["*"]}, \texttt{allow\_credentials=True}, \texttt{allow\_methods=["*"]}, \texttt{allow\_headers=["*"]}.

\textbf{API Documentation:} FastAPI automatically generates interactive documentation accessible at \texttt{/docs} (Swagger UI) and \texttt{/redoc} (ReDoc). Developers can test endpoints directly through browser interface with example request bodies and responses.

\section{System Deployment Guide}
This section provides step-by-step instructions for TA to set up and run the system.

\subsection{Database Setup}
\textbf{Step 1: Create Database}
\begin{lstlisting}[style=sqlstyle]
CREATE DATABASE dormitory_management_system 
CHARACTER SET utf8mb4 COLLATE utf8mb4_unicode_ci;
USE dormitory_management_system;
\end{lstlisting}

\textbf{Step 2: Import SQL Scripts} \\
Navigate to the \texttt{sql/} folder and execute the SQL files in order:
\begin{lstlisting}[style=sqlstyle]
mysql -u root -p dormitory_management_system < sql/01_create_tables.sql
mysql -u root -p dormitory_management_system < sql/02_insert_data.sql
\end{lstlisting}

The database will be populated with 4,354 students, 1,350 dormitories, sample bills, and request records. Most data are real-world data aquired from SYSU. Thank bakabaka9405 for providing the dataset.

\subsection{Backend Setup and Launch}
\textbf{Step 1: Install Dependencies}
\begin{lstlisting}[language=bash, basicstyle=\ttfamily\scriptsize]
cd backend
pip install -r requirements.txt
\end{lstlisting}

\textbf{Step 2: Configure Database Connection} \\
Edit \texttt{backend/app/database.py} to set MySQL credentials:
\begin{lstlisting}[language=Python, basicstyle=\ttfamily\scriptsize]
DATABASE_URL = "mysql+pymysql://root:password@localhost:3306/dormitory_management_system"
\end{lstlisting}

\textbf{Step 3: Start Backend Server}
\begin{lstlisting}[language=bash, basicstyle=\ttfamily\scriptsize]
uvicorn app.main:app --reload --host 0.0.0.0 --port 8000
\end{lstlisting}
Backend will run at \texttt{http://localhost:8000}. API documentation available at \texttt{http://localhost:8000/docs}.

\subsection{Frontend Setup and Launch}
\textbf{Step 1: Install Dependencies}
\begin{lstlisting}[language=bash, basicstyle=\ttfamily\scriptsize]
cd frontend
npm install
\end{lstlisting}

\textbf{Step 2: Start Development Server}
\begin{lstlisting}[language=bash, basicstyle=\ttfamily\scriptsize]
npm run dev
\end{lstlisting}
Frontend will run at \texttt{http://localhost:5173}.

\subsection{Test Accounts}
\textbf{Student Accounts:} All students use default password \texttt{123456}. Login with any student ID from the database (e.g., \texttt{123090669}, \texttt{123090123}).

\textbf{Administrator Accounts:}
\begin{itemize}
\item Username: \texttt{admin}, Password: \texttt{admin123}
\item Username: \texttt{dorm\_manager}, Password: \texttt{manager123}
\item Username: \texttt{maintenance}, Password: \texttt{maint123}
\end{itemize}

\section{Summary and Harvest}
\subsection{Achievements}
Successfully implemented a production-ready system with 21+ functions across 2 roles, frontend-backend separation, normalized 6-table database, Argon2+JWT security, and responsive web interface.

\subsection{Skills Acquired}
\textbf{Backend:} Mastered FastAPI async/await patterns, SQLAlchemy relationships and query optimization, JWT authentication, RESTful API design, and database transaction management with ACID properties.

\textbf{Frontend:} Proficient in Vue 3 Composition API (\texttt{ref}, \texttt{reactive}, \texttt{computed}), Vue Router with navigation guards, Element Plus components, responsive layouts (CSS Grid/Flexbox), and asynchronous state management.

\textbf{Database:} Applied normalization (3NF), designed ER models, wrote complex SQL (joins, aggregates, subqueries), optimized with indexing, and implemented atomic transactions.

\textbf{Engineering:} Adopted layered architecture, Git version control, comprehensive error handling, API documentation (Swagger), and testing (unit, integration, UAT).

\textbf{Security:} Implemented parameterized queries (SQL injection prevention), Argon2 password hashing, HTTPS encryption, authentication/authorization, and least privilege principle.

\subsection{Key Challenges and Solutions}
Migrated to Argon2 password hashing for security; implemented atomic transactions with rollback for data consistency; centralized authentication with localStorage; standardized error handling using Axios interceptors.

\subsection{Conclusion}
This project provided hands-on full-stack development experience, demonstrating how database, backend API, and frontend UI integrate cohesively. The modular architecture with clear interfaces ensures maintainability and scalability, preparing me for future software engineering challenges.

\end{document}
